---
title: 'New York Vehicle Weight Distribution Analysis'
subtitle: 'VIN-Based Fleet Weight Distribution by Model Year'
author: 'Michael McLean | mclean.bsky.social'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: false
    theme: flatly
    css: |
      body { font-family: 'Segoe UI', Tahoma, Geneva, Verdana, sans-serif; }
      .main-container { max-width: 1200px; }
      h1, h2, h3 { color: #2E86AB; }
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Load required libraries
library(httr2)
library(jsonlite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

# Set global options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  fig.align = "center"
)

# Define EPA-style weight bins (matching the figure you showed)
weight_bins <- c(0, 2750, 3000, 3500, 4000, 4500, 5000, 5500, 6000, Inf)
weight_labels <- c("<2750", "2750-3000", "3000-3500", "3500-4000", "4000-4500", "4500-5000", "5000-5500", "5500-6000", "≥6000")
```

# New York Vehicle Weight Distribution

This analysis examines the weight distribution of registered passenger
vehicles in New York State using VIN-level data from the NY DMV
registration database. The data is filtered to passenger vehicles only
(VEH record type) and shows how the fleet weight distribution has
evolved over time.

```{r data-fetch, echo=false}
# Function to fetch NY vehicle data with pagination
fetch_ny_vehicle_data <- function(max_records = NULL, batch_size = 50000) {
  
  cat("Fetching New York vehicle registration data with pagination...\n")
  
  # Base URL for individual vehicle records (not pre-aggregated)
  base_url <- "https://data.ny.gov/resource/w4pv-hbkt.json"
  
  # Initialize variables for pagination
  all_data <- list()
  offset <- 0
  batch_num <- 1
  total_fetched <- 0
  
  repeat {
    cat("Fetching batch", batch_num, "(offset:", scales::comma(offset), ")...\n")
    
    # Build query parameters
    query_params <- list(
      `$where` = "record_type = 'VEH'",  # Filter to passenger vehicles only
      `$select` = "model_year,unladen_weight,vin",  # Select only needed columns
      `$limit` = batch_size,
      `$offset` = offset,
      `$order` = "model_year"  # Order for consistent pagination
    )
    
    # Make the API request
    tryCatch({
      response <- request(base_url) %>%
        req_url_query(!!!query_params) %>%
        req_perform()
      
      # Parse JSON response
      batch_data <- response %>%
        resp_body_json(simplifyVector = TRUE)
      
      # Check if we got any data
      if (length(batch_data) == 0 || nrow(as_tibble(batch_data)) == 0) {
        cat("No more data available. Stopping pagination.\n")
        break
      }
      
      # Convert to tibble and store
      batch_tibble <- as_tibble(batch_data)
      all_data[[batch_num]] <- batch_tibble
      
      batch_records <- nrow(batch_tibble)
      total_fetched <- total_fetched + batch_records
      
      cat("Batch", batch_num, "completed:", scales::comma(batch_records), "records\n")
      cat("Total fetched so far:", scales::comma(total_fetched), "records\n")
      
      # Check if we've reached the maximum records limit
      if (!is.null(max_records) && total_fetched >= max_records) {
        cat("Reached maximum records limit (", scales::comma(max_records), "). Stopping.\n")
        break
      }
      
      # Check if we got fewer records than requested (end of data)
      if (batch_records < batch_size) {
        cat("Received fewer records than requested. End of data reached.\n")
        break
      }
      
      # Update for next iteration
      offset <- offset + batch_size
      batch_num <- batch_num + 1
      
      # Add a small delay to be respectful to the API
      Sys.sleep(0.5)
      
    }, error = function(e) {
      cat("Error fetching batch", batch_num, ":", e$message, "\n")
      break
    })
  }
  
  # Combine all batches
  if (length(all_data) > 0) {
    combined_data <- bind_rows(all_data)
    cat("\nData fetching completed!\n")
    cat("Total batches:", length(all_data), "\n")
    cat("Total records:", scales::comma(nrow(combined_data)), "\n")
    cat("Unique VINs:", scales::comma(n_distinct(combined_data$vin)), "\n")
    
    return(combined_data)
  } else {
    cat("No data was fetched.\n")
    return(tibble())
  }
}

# Fetch the data (limit to 200k records for initial analysis)
ny_raw_data <- fetch_ny_vehicle_data(max_records = 13000000, batch_size = 50000)
```

```{r data-processing, echo=false}
# Clean and process the data
ny_processed <- ny_raw_data %>%
  # Convert columns to appropriate types
  mutate(
    model_year = as.numeric(model_year),
    unladen_weight = as.numeric(unladen_weight)
  ) %>%
  # Filter for reasonable model years and weights
  filter(
    unladen_weight > 0,
    !is.na(model_year) & !is.na(unladen_weight) & !is.na(vin)
  ) %>%
  # Remove duplicate VINs (keep first occurrence)
  distinct(vin, .keep_all = TRUE) %>%
  # Create EPA-style weight bins
  mutate(
    weight_class = cut(unladen_weight, 
                      breaks = weight_bins, 
                      labels = weight_labels, 
                      right = FALSE, 
                      include.lowest = TRUE)
  ) %>%
  # Remove any rows where weight_class couldn't be assigned
  filter(!is.na(weight_class))

# Calculate shares by model year and weight class
weight_distribution <- ny_processed %>%
  group_by(model_year, weight_class) %>%
  summarise(
    vehicles = n(),  # Count individual vehicles instead of summing pre-aggregated counts
    .groups = "drop"
  ) %>%
  group_by(model_year) %>%
  mutate(
    year_total = sum(vehicles),
    share = vehicles / year_total
  ) %>%
  ungroup() %>%
  # Complete the data to ensure all weight classes are present for each year
  complete(model_year, weight_class, fill = list(vehicles = 0, share = 0)) %>%
  group_by(model_year) %>%
  mutate(
    year_total = sum(vehicles),
    share = if_else(year_total > 0, vehicles / year_total, 0)
  ) %>%
  ungroup()
```

## Weight Class Distribution by Model Year

This chart replicates the EPA's Figure 3.6 format, showing how New
York's passenger vehicle fleet weight distribution has changed over
time. Each colored band represents a different weight class, with the
height showing that class's share of total vehicle registrations for
each model year.

```{r stacked-area-plot, echo=false}
# Create the stacked area plot matching EPA Figure 3.6 style
weight_distribution_chart <- weight_distribution %>% 
  filter(model_year >= 1970) %>%
  ggplot(aes(x = model_year, y = share, fill = weight_class)) +
  geom_area(position = "stack", alpha = 0.9) +
  scale_fill_manual(
    name = "Weight (lb)",
    values = c(
      "<2750" = "#1f4e79",           # Dark blue
      "2750-3000" = "#2e5f8a",       # Medium dark blue  
      "3000-3500" = "#4472a8",       # Medium blue
      "3500-4000" = "#6ba6cd",       # Light blue
      "4000-4500" = "#a2c4e0",       # Very light blue
      "4500-5000" = "#f4e4a6",       # Light yellow
      "5000-5500" = "#f2cc8f",       # Light orange
      "5500-6000" = "#e07b39",       # Orange
      "≥6000" = "#d73027"            # Red
    )
  ) +
  scale_x_continuous(
    breaks = seq(1970, 2025, 5),
    limits = c(1970, 2025)
  ) +
  scale_y_continuous(
    labels = percent_format(),
    breaks = seq(0, 1, 0.25),
    limits = c(0, 1)
  ) +
  labs(
    title = "Weight Class Distribution by Model Year",
    subtitle = "New York State Passenger Vehicle Registrations",
    x = "Model Year",
    y = "Share of Active Passenger Vehicle Registrations",
    caption = "Data: NY DMV Vehicle Registration Database | VEH records only"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold", color = "#2E86AB"),
    plot.subtitle = element_text(size = 14, color = "#457B9D"),
    axis.title = element_text(color = "#457B9D"),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray95", size = 0.5),
    plot.caption = element_text(size = 10, color = "gray60")
  ) +
  guides(fill = guide_legend(reverse = TRUE, ncol = 1))
```

## Summary Statistics

```{r summary-stats, echo=false}
library(dplyr)
library(kableExtra)
library(scales)

# Create the summary with proper ordering and total row
weight_summary_formatted <- weight_summary %>%
  # Create a sorting key for proper weight class ordering
  mutate(
    sort_order = case_when(
      weight_class == "<2750" ~ 1,
      weight_class == "2750-3000" ~ 2,
      weight_class == "3000-3500" ~ 3,
      weight_class == "3500-4000" ~ 4,
      weight_class == "4000-4500" ~ 5,
      weight_class == "4500-5000" ~ 6,
      weight_class == "5000-5500" ~ 7,
      weight_class == "5500-6000" ~ 8,
      weight_class == "≥6000" ~ 9,
      TRUE ~ 10
    )
  ) %>%
  arrange(sort_order) %>%
  select(-sort_order) %>%
  # Add total row
  bind_rows(
    summarise(., 
              weight_class = "Total",
              total_vehicles = sum(total_vehicles),
              share_of_total = 1.0)
  ) %>%
  # Format the weight classes with commas
  mutate(
    weight_class_formatted = case_when(
      weight_class == "<2750" ~ "<2,750",
      weight_class == "2750-3000" ~ "2,750-3,000", 
      weight_class == "3000-3500" ~ "3,000-3,500",
      weight_class == "3500-4000" ~ "3,500-4,000",
      weight_class == "4000-4500" ~ "4,000-4,500",
      weight_class == "4500-5000" ~ "4,500-5,000",
      weight_class == "5000-5500" ~ "5,000-5,500",
      weight_class == "5500-6000" ~ "5,500-6,000",
      weight_class == "≥6000" ~ "≥6,000",
      TRUE ~ weight_class
    ),
    total_vehicles = comma(total_vehicles),
    share_of_total = percent(share_of_total, accuracy = 0.1)
  ) %>%
  select(weight_class_formatted, total_vehicles, share_of_total)

# Create the kable
weight_summary_formatted %>%
  kable(
    col.names = c("Weight Class (lbs)", "Total Vehicles", "Share of Total"),
    caption = "Vehicle Distribution by Weight Class",
    align = c("l", "r", "r"),
    escape = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2c3e50") %>%
  row_spec(nrow(weight_summary_formatted), bold = TRUE, background = "#f8f9fa") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(
    general = "Based on vehicle registration data across all years in dataset.",
    general_title = "",
    footnote_as_chunk = TRUE
  )
```

------------------------------------------------------------------------

*Data source: New York State DMV Vehicle Registration Database via NY
[Open Data
Portal](https://data.ny.gov/Transportation/Vehicle-Snowmobile-and-Boat-Registrations/w4pv-hbkt/about_data).
Analysis limited to passenger vehicles (VEH record type) only.*
